Reinforcement Learning Produces Dominant Strategies for the Iterated Prisoner's Dilemma

### 2017-07-05 main.tex
Run analysis with 25k repetitions.
Some minor esthetic improvements.
- Increase the font size of labels
- Increase the height for landscape images.
- Rearrange group citations
- Fix probabilistic notation

There are several previous publications that use evolutionary algorithms to
evolve IPD strategies in various circumstances \cite{marks1989niche} \cite{fogel1993evolving} \cite{ashlock2006training}
\cite{ashlock2006changes} \cite{vassiliades2010multiagent} \cite{ashlock2014shaped}
\cite{ashlock2014evolution} \cite{ashlock2015multiple} \cite{barlow2015varying}
\cite{sudo2015effects}. See also \cite{gaudesi2016exploiting} for a strategy
trained to win against a collection of well-known IPD opponents and see
\cite{franken2005particle} for a prior use of particle swarm algorithms. Our
results are unique in that we are able to train against a large collection of
well-known strategies available in the scientific literature.

@@

The Axelrod library now contains many strategies based on machine learning
methods. Most are deterministic, use many rounds of memory, and perform
extremely well in tournaments.

@@

The first strategy trained with reinforcement learning in the library is based
on lookup tables. The strategy encodes a set of deterministic responses
based on the opponent's first $n_1$ moves, the opponent's last $m_1$ moves, and
the players last $m_2$ moves. If $n_1 > 0$ then the player has infinite memory
depth, otherwise it has depth $\max{m_1, m_2}$. Although we tried various
combinations of $n_1, m_1,$ and $m_2$, the best performance at the time of
training was obtained for $n_1 = m_1 = m_2 = 2$ and generally for $n_1 > 0$.
First impressions matter in the IPD. The library includes a strategy
called EvolvedLookerUp2\_2\_2 which is among the top strategies in the library.

@@

This strategy type can be used to train arbitrary memory-$n$ strategies and a
memory one strategy was trained and is called PSO Gambler Mem 1, with
probabilities $(\text{Pr(C|CC)}, \text{Pr(C|CD)}, \text{Pr(C|DC)}, \text{Pr(C|DD)}) = (1, 0.5217, 0, 0.121)$. Though it performs well in deterministic tournaments
it is not as good as the longer memory strategies, and is bested by a similar
strategy that also uses the first round of play.

@@

\begin{landscape}
    \begin{figure}[!hbtp]
        \includegraphics[width=\paperwidth]{./assets/standard_scores_boxplots.pdf}
        \caption{Standard Tournament: Mean score per turn (ranked by median
        over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments)}
        \label{fig:standard_boxplot}
    \end{figure}
\end{landscape}

@@

\begin{figure}[!hbtp]
    \includegraphics[width=\textwidth]{./assets/standard_scores_heatmap.pdf}
    \caption{Standard Tournament: Mean score per turn of row players against
    column players (ordered by median over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments)}
    \label{fig:standard_heatmap}
\end{figure}

@@

\begin{landscape}
    \begin{figure}[!hbtp]
        \centering
        \includegraphics[width=\textwidth]{./assets/standard_wins_boxplots.pdf}
        \caption{Standard Tournament: number of wins per tournament (ordered by
        median over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments)}
        \label{fig:standard_winplot}
    \end{figure}
\end{landscape}

@@

\begin{landscape}
    \begin{figure}[!hbtp]
        \centering
        \includegraphics[width=\textwidth]{./assets/standard_ranks_boxplots.pdf}
        \caption{Standard Tournament: rank in each tournament (ordered by
        median over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments)}
        \label{fig:standard_ranks_boxplot}
    \end{figure}
\end{landscape}

@@

\begin{landscape}
    \begin{figure}[!hbtp]
        \includegraphics[width=\paperwidth]{./assets/noisy_scores_boxplots.pdf}
        \caption{Noisy (5\%) Tournament: Mean score per turn (ranked by median
        over
        \protect\input{./assets/noisy_number_of_repetitions.tex}tournaments)}
        \label{fig:noisy_score}
    \end{figure}
\end{landscape}

@@

\begin{figure}[!hbtp]
    \includegraphics[width=\textwidth]{./assets/noisy_scores_heatmap.pdf}
    \caption{Noisy (5\%) Tournament: Mean score per turn of row players against
    column players (ordered by median over
        \protect\input{./assets/noisy_number_of_repetitions.tex}tournaments)}
    \label{fig:noisy_heatmap}
\end{figure}

@@

\begin{landscape}
    \begin{figure}[!hbtp]
        \centering
        \includegraphics[width=\textwidth]{./assets/noisy_wins_boxplots.pdf}
        \caption{Noisy (5\%) Tournament: number of wins per tournament (ordered by
        median over
        \protect\input{./assets/noisy_number_of_repetitions.tex}tournaments)}
        \label{fig:noisy_winplot}
    \end{figure}
\end{landscape}

@@

Finally, Table~\ref{tbl:noisy_ranks} and
Figure~\ref{fig:noisy_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. We see, that the stochasticity
of the ranks understandably increases the DBS strategy never ranks lower than
third and wins 75\% of the time. The two strategies trained for noisy
tournaments rank in the top three 95\% of the time.
% TODO Possibly update this with more data.

@@

\begin{landscape}
    \begin{figure}[!hbtp]
        \centering
        \includegraphics[width=\textwidth]{./assets/noisy_ranks_boxplots.pdf}
        \caption{Noisy (5\%) Tournament: rank in each tournament (ordered by
        median over
        \protect\input{./assets/noisy_number_of_repetitions.tex}tournaments)}
        \label{fig:noisy_ranks_boxplot}
    \end{figure}
\end{landscape}

>>>>

There are several previous publications that use evolutionary algorithms to
evolve IPD strategies in various circumstances
\cite{ashlock2006training, ashlock2015multiple, ashlock2006changes,
      ashlock2014shaped, ashlock2014evolution, barlow2015varying,
      fogel1993evolving, marks1989niche, sudo2015effects,
      vassiliades2010multiagent}. See also \cite{gaudesi2016exploiting} for a
strategy trained to win against a collection of well-known IPD opponents and see
\cite{franken2005particle} for a prior use of particle swarm algorithms. Our
results are unique in that we are able to train against a large collection of
well-known strategies available in the scientific literature.

@@

The Axelrod library now contains many parametrised strategies trained using
machine learning
methods. Most are deterministic, use many rounds of memory, and perform
extremely well in tournaments.

The various archetypes will be described in the following sections.

@@

The first strategy trained with reinforcement learning in the library is based
on lookup tables. The strategy encodes a set of deterministic responses
based on the opponent's first $n_1$ moves, the opponent's last $m_1$ moves, and
the players last $m_2$ moves. If $n_1 > 0$ then the player has infinite memory
depth, otherwise it has depth $\max{m_1, m_2}$. Although we tried various
combinations of $n_1, m_1,$ and $m_2$, the best performance at the time of
training was obtained for $n_1 = m_1 = m_2 = 2$ and generally for $n_1 > 0$.
First impressions matter in the IPD\@. The library includes a strategy
called EvolvedLookerUp2\_2\_2 which is among the top strategies in the library.

@@

This strategy type can be used to train arbitrary memory-$n$ strategies and a
memory one strategy was trained and is called PSO Gambler Mem 1, with
probabilities $(\text{Pr}(\text{C}\;|\;\text{CC}),
                \text{Pr}(\text{C}\;|\;\text{CD}),
                \text{Pr}(\text{C}\;|\;\text{DC}),
                \text{Pr}(\text{C}\;|\;\text{DD})) = (1, 0.5217, 0, 0.121)$.
Though it performs well in standard tournaments (see
Table~\ref{tbl:standard_score})
it is not as good as the longer memory strategies, and is bested by a similar
strategy that also uses the first round of play: PSO Gamble 1 1 1.

@@

\begin{landscape}
    \begin{figure}[!hbtp]
        \centering
        \includegraphics[width=\paperwidth]{./assets/standard_scores_boxplots.pdf}
        \caption{Standard Tournament: Mean score per turn (ranked by median
        over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments)}
        \label{fig:standard_boxplot}
    \end{figure}
\end{landscape}

@@

\begin{figure}[!hbtp]
    \centering
    \includegraphics[width=\textwidth]{./assets/standard_scores_heatmap.pdf}
    \caption{Standard Tournament: Mean score per turn of row players against
    column players (ranked by median over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments)}
    \label{fig:standard_heatmap}
\end{figure}

@@

\begin{landscape}
    \begin{figure}[!hbtp]
        \centering
        \includegraphics[width=\paperwidth]{./assets/standard_wins_boxplots.pdf}
        \caption{Standard Tournament: number of wins per tournament (ranked by
        median over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments)}
        \label{fig:standard_winplot}
    \end{figure}
\end{landscape}

@@

\begin{landscape}
    \begin{figure}[!hbtp]
        \centering
        \includegraphics[width=\paperwidth]{./assets/standard_ranks_boxplots.pdf}
        \caption{Standard Tournament: rank in each tournament (ranked by
        median over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments)}
        \label{fig:standard_ranks_boxplot}
    \end{figure}
\end{landscape}

@@

\begin{landscape}
    \begin{figure}[!hbtp]
        \centering
        \includegraphics[width=\paperwidth]{./assets/noisy_scores_boxplots.pdf}
        \caption{Noisy (5\%) Tournament: Mean score per turn (ranked by median
        over
        \protect\input{./assets/noisy_number_of_repetitions.tex}tournaments)}
        \label{fig:noisy_score}
    \end{figure}
\end{landscape}

@@

\begin{figure}[!hbtp]
    \centering
    \includegraphics[width=\textwidth]{./assets/noisy_scores_heatmap.pdf}
    \caption{Noisy (5\%) Tournament: Mean score per turn of row players against
    column players (ranked by median over
        \protect\input{./assets/noisy_number_of_repetitions.tex}tournaments)}
    \label{fig:noisy_heatmap}
\end{figure}

@@

\begin{landscape}
    \begin{figure}[!hbtp]
        \centering
        \includegraphics[width=\paperwidth]{./assets/noisy_wins_boxplots.pdf}
        \caption{Noisy (5\%) Tournament: number of wins per tournament (ranked by
        median over
        \protect\input{./assets/noisy_number_of_repetitions.tex}tournaments)}
        \label{fig:noisy_winplot}
    \end{figure}
\end{landscape}

@@

Finally, Table~\ref{tbl:noisy_ranks} and
Figure~\ref{fig:noisy_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. We see, that the stochasticity
of the ranks understandably increases the DBS strategy never ranks lower than
second and wins 75\% of the time. The two strategies trained for noisy
tournaments rank in the top three 95\% of the time.
% TODO Possibly update this with more data.

@@

\begin{landscape}
    \begin{figure}[!hbtp]
        \centering
        \includegraphics[width=\paperwidth]{./assets/noisy_ranks_boxplots.pdf}
        \caption{Noisy (5\%) Tournament: rank in each tournament (ranked by
        median over
        \protect\input{./assets/noisy_number_of_repetitions.tex}tournaments)}
        \label{fig:noisy_ranks_boxplot}
    \end{figure}
\end{landscape}
